A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-11 code/figure4.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

layout <- "ABBCDD"
dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")

```

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 4: Gene level analysis reveals tissue-specific class signatures.** Odds ratio plot; Column uses a logarithmic axis to represent odds ratio with a 95% confidence interval. Row represents the significant genes from each cluster. Each cluster was compared against the background (all other clusters) to find deferentially mutated genes. Significant genes (pvalue < 0.01) were selected and limited to the top ten results for each clusters. Heatmap; Column represents the cancer type. Row represents the significant genes as described above. The heatmap shows the proportion of samples per cluster, per cancer type that is mutated for each of the significant genes. Depleted significant genes (left). Enriched significant genes (right).</p>

>>>>

layout <- "ABBCDD"
dn_p + tdat_dn_p + up_p + tdat_up_p + plot_layout(guides = 'collect', design = layout) & theme(legend.position = "bottom")
```

### 2020-08-11 code/supplements.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

### Supplementary Figure S1

```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
#out.width = '90%',
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
![**Supplementary figure S1. Distribution of cancer type on UMAP clusters.** Each panel shows the identity of individual tumors by one of 23 cancer types projected onto the cluster UMAP identified in **Figure 1 A** of the main text._]

### Supplementary Figure S2

```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```

![**Supplementary Figure S2. Histological subtypes.** _Each panel shows the histological subtypes of one of the 23 cancers projected on to the tumors of the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S3

```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
```

![**Supplementary Figure S3. Common oncogenic mutations.** _Each panel shows one of 20 commonly mutated oncogenic genes projected on the UMAP from **Figure 1A of the main text.**_]

### Supplementary Figure S4

```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```

![**Supplementary Figure S4. Projection of pathway disruptions onto the UMAP.**_]

```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
```

### Supplementary Table ST1

```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

>>>>

### Supplementary Figure S1
```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
#out.width = '90%',
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
\newpage
### Supplementary Figure S2
```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```
\newpage
### Supplementary Figure S3
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
```
\newpage
### Supplementary Figure S4
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```
\newpage
### Supplementary Figure S5
```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
```
\newpage
### Supplementary Table 1
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

### 2020-08-11 code/table2.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

```{r table-2, eval = FALSE}
## this is for html only
library(kableExtra)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, dpi=600)
load("../data/m1_minus_m0_per_cluster_rbeta.Rda")

@@

```{r, echo = FALSE}
## docx friendly
load("../data/m1_minus_m0_per_cluster_rbeta.Rda")

@@

table2 %>% flextable() %>% 
  set_caption(caption = "") %>%
  align(part = "all") %>% 
  width(., width = c(4,0.75,1,1)) %>% 
  fontsize(size = 10, part = "body")

```

>>>>

```{r table-2, eval = FALSE}
## this is for html only
load("../data/m1_minus_m0_per_cluster_rbeta.Rda")

@@

```{r table2-docx-friendly}
## docx friendly
load("../data/m1_minus_m0_per_cluster_rbeta.Rda")

@@

table2 <-table2 %>% flextable() %>% 
   set_caption(caption = "") %>%
   align(part = "all") %>% 
   width(., width = c(4,0.75,1,1)) %>% 
   fontsize(size = 10, part = "body")

knitr::knit_print(table2)
```

### 2020-08-11 text/manuscript.Rmd
Add \newpage to separate plots, fix knit error with the word-style, update text for figure 3

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
  Dennis J. Hazelett$^{1,3}$
bibliography: references.bib
output:
  word_document:
    reference_docx: word-style-reference-ttmanu.docx
always_allow_html: TRUE
editor_options:
  chunk_output_type: console
---

@@

### A taxonomy of tumors based on disrupted molecular pathways

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 8940 genes; see methods for details).

@@

As discussed previously, prior efforts to extract common signatures from
pan-cancer datasets met with difficulty in distinguishing tumor samples from
their tissue-specific -omics data signatures. Given our pathway-disruption based
classification, this raises the question, are tumor phenotypes driven entirely
by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
few samples to detect above statistical significance thresholds), or a
combination of both? To answer this question, we compared each cluster relative
to the background to find deferentially mutated genes. We ranked the odd ratios 
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each 
cluster (**Figure 4**; odds ratios plot). We find that the clusters possess 
unique gene sets. Cluster 2 is enriched for genes responsible for protein
homeostasis. Cluster 4 is enriched for genes responsible for nucleosome 
structure. Clusters 7 and 8, which shared multiple enrichment in signaling
pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9 is enriched for both _PIK3CA_ and _KRAS_ which may
suggest tumors in this cluster derived from either clusters 7 or 8. Next, we 
investigated the proportion of samples per cancer type for the significant genes 
within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors 
were not predominated by one or more highly mutated genes across all cancers. 
Instead, some cancer-specific tumors in these classes had mutation profiles that
differ from one another (_e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC
was enriched for _UBC_, and PAAD was enriched for _PSMA1_).

@@

```{r child='survival-methods.Rmd'}
```

```{r child='../code/figure1.Rmd'}
```

```{r child='../code/figure2.Rmd'}
```

```{r child = '../code/figure4.Rmd'}
```

```{r child='../code/table2.Rmd'}
```

![](../figures/fig6B.png)
![](../figures/fig6C.png)

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and
  Dennis J. Hazelett$^{1,3}$
bibliography: references.bib
output:
  word_document:
    reference_docx: "../text/word-style-reference-ttmanu.docx"
always_allow_html: TRUE
editor_options:
  chunk_output_type: console
---

@@

### A taxonomy of tumors based on disrupted molecular pathways
To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data because were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to start with somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we focused on all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 8940 genes; see methods for details).

@@

As discussed previously, prior efforts to extract common signatures from pan-cancer 
datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
-omics data signatures. Given our pathway-disruption based classification, this raises 
the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
thresholds), or a combination of both? To answer this question, we compared each cluster 
relative to the background to find deferentially mutated genes. We ranked the odd ratios
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
(**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
that tumors in this cluster could be a derivative either cluster 7 or 8. Clusters 3 and 
5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
Next, we investigated the proportion of samples per cancer type for the significant genes
within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors were not 
predominated by one or more highly mutated genes across all cancers. Instead, several 
cancer-specific tumors in these classes had mutation profiles that differ from one another 
( _e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC was enriched for _UBC_, and 
PAAD was enriched for _PSMA1_). Taken together, our data identifies a framework cancer 
type specific mutation associated with a specific cluster.

@@

```{r child='survival-methods.Rmd'}
```
\newpage
```{r child='../text/fig_legend.Rmd'}
```

\newpage
```{r child='../code/figure1.Rmd'}
```
\newpage
```{r child='../code/figure2.Rmd'}
```
\newpage
```{r child = '../code/figure4.Rmd'}
```
\newpage
```{r child='../code/table2.Rmd'}
```
\newpage

![](../figures/fig6B.png)
\pagebreak
![](../figures/fig6C.png)
